Future challenges for Research Software Engineering

### 2025-09-22 future_challenges_for_RSEng.tex
spell check

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\author{
Florian Goth\authorOrcid{0000-0003-2707-4790}\authorRef{1},
Leyla Jael Castro\authorOrcid{0000-0003-3986-0510}\authorRef{2},
Gerasimos Chourdakis\authorOrcid{0000-0002-3977-1385}\authorRef{3},
\texorpdfstring{\\}{} % line break
Simon Christ\authorOrcid{0000-0002-5866-1472}\authorRef{4},
Jeremy Cohen\authorOrcid{0000-0003-4312-2537}\authorRef{5},
Nicholas A. Del Grosso\thinspace\authorRef{6},
\texorpdfstring{\\}{} % line break
Jean-Noël Grad\authorOrcid{0000-0002-5821-4912}\authorRef{7},
Magnus Hagdorn\authorOrcid{0000-0002-5076-4864}\authorRef{8},
Toby Hodges\authorOrcid{0000-0003-1766-456X}\authorRef{9},
Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10},
\texorpdfstring{\\}{} % line break
Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{11},
Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{12},
Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{13}
} % Authors and references to addresses
\institute{% Institutes with labels
\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\\
\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\\
\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\\
\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\\
\autlabel{5} Imperial College London, London, UK\\
\autlabel{6} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany\\
\autlabel{7} Institute for Computational Physics, University of Stuttgart, Germany\\
\autlabel{8} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\\
\autlabel{9} The Carpentries, USA\\
\autlabel{10} Technische Universität Braunschweig, Germany\\
\autlabel{11} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\\
\autlabel{12} Independent researcher, Germany\\
\autlabel{13} Weierstrass Institute, Berlin, Germany;
              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\\
}
\abstract{
It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
As the field develops, the role and value of Research Software Engineers (RSEs) becomes clearer.
With the growing importance and scope of software engineering in research,
researchers who code now begin to fill dedicated, specialized RSE roles crucial for the sustainable development
of long-term research software projects that can rapidly innovate.
Looking into the future, participants of the deRSE25 conference discussed how RSEng could evolve in response to the
rapid changes in the broader social, technological, and organisational landscape.
This paper presents the main discussion points of the conference participants,
imagining a future in which RSEs have a well-defined and well-integrated role in research institutions,
working as engineers of ever more complex research software projects,
taming a growing range of methods and tools,
and guiding researchers through technical and ethical considerations
related to research software.
} % Abstract of the article
\keywords{Future RSE, Evolving field, Vision} % Keywords for the article

@@

\begin{story}{The Definition of RSEs}{defrse}
Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
Now, their roles are unrecognizable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who transformed the chaos back into meaningful order.
They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
Their job titles have changed, but their role as bridges between logic and life remains the same.
In a future of shifting code and constant reinvention, it is their understanding of people that never went out of date.

@@

Further risks that were later discussed include an intensified replication crisis,
a de-democratization of research, and loss of trust in science.
Due to the non-deterministic nature of generative AI, we expect AI-assisted simulations
and LLM-assisted data analysis to become less reproducible.
This is not an entirely new issue, and similar concerns exist in the HPC community,
where reproducing simulation data and data analysis workflows presents tough challenges~\cite{Antunes2024}.
Professional RSEs could safeguard reproducibility by designing, evaluating, and monitoring appropriate workflows.
When applicable, standardized and version-controlled models should be preferred
over self-developed models, and self-developed models should be treated like any
other research output (i.e. publish the trained model, training data and training workflow).
Future research will require significant computing and storage resources, as well as useful data and models,
therefore it will be more likely to be driven by institutions that are able to provide these.
Finally, an increasing risk is the loss of trust in science,
in particular due to the automation of the peer-reviewing process~\cite{Naddaf2025}
and the malicious use of hidden prompts~\cite{Gibney2025}.
As researchers increase their reliance on AI assistants,
RSEs will be freed from repeatedly crafting similar individual elements,
but will be spending more of their time designing and reviewing software written by other humans and AI tools.
To address these risks, the RSE of the future should not only steer every part
of the research software development cycle (ensuring dissemination, archiving, and reproducibility),
but also contribute to endangered community resources and infrastructure that
will keep research accessible to the wider scientific community.

@@

% society implications
The work being undertaken by RSEs may well have direct, or indirect,
ethical implications in areas such as social engineering~\cite{s2erc,Siadati2024},
climate~\cite{Lannelongue2023} and more.
Like all areas of research and engineering RSEs are faced with the dual use dilemma~\cite{Bobier2024} where artefacts they develop can be used for military purposes.
As engineers, RSEs take the responsibility for the software they develop.
This includes an awareness of ethical issues that may arise.
RSEs need to be aware of these issues and develop their own ``red lines''
which need to be supported by organizational and professional guidelines.
With the advantages of \autoref{sec:ai}, ever-more AI created artefacts will be included in research.
As a result of this, we expect a discussion within societies about their acceptable use.

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{}{academic}
We expect research software to be a central part of scientific practice, on the same level as data and publications.
This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specializing in certain tasks like data management, HPC, or visualisation.
At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
Additionally, researchers will seamlessly interact with infrastructures for using research software.
Replacement of staff reaching retirement has led to the creation of more RSE professorships.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of societal generations.
While sounding negative, this can also be an opportunity,
if in the time-frame till 2035 a large number of aging, permanent staff is replaced by new people that have observed the
digital transformation first-hand and have actively participated in it through various grassroots initiatives.\\
On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

@@

Besides all the improvements, the workshop participants still expect some challenges to persist,
and new challenges to emerge.
Due to the rise of dedicated RSE units, networking people and teams,
federating infrastructures, as well as organizing and sharing knowledge will become important.
Structuring these new units and joining their administrative support will also help the units to grow.
A particular example is hiring:
While a standardised RSE Master's degree will make it easier to develop the next generation of RSEs,
organizations could develop standardized hiring processes to acquire the skill set that they actually need.
Budgeting and administrative hurdles will also probably still exist and hinder the formation of permanent and well-funded RSE teams.
However, we expect the growing (perceived) importance of research software
to convince research councils and funders to provide reliable streams of funding.

@@

\begin{story}{New career paths}{New career paths}
Ken, though trained as a physicist, found his passion not in theory,
but in keeping the quantum computers humming. He became the go-to expert
for troubleshooting entangled qubits and stabilizing delicate cryogenic systems.
His unique blend of physics knowledge and hands-on problem solving earned him
a new title among his peers: the first Quantum RTP.
In a world where theory meets reality, Ken quietly ensures that the future
of quantum computing stays on track.
\end{story}
\section{What is in this future for RSEs?}
%FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
%GC-NOTE: Shall we add another story box here? Ideas what we write here?
In the previous sections, we have tried to list some of the trends that we foresee.
The material presented here is based on the outputs of a workshop, and extended with the authors' ideas.
Nevertheless, there are some things that we declared out of scope here.
Among that is the ongoing digital transformation in wider society.
Digital competencies in the general population continue to increase,
making the jump to software development easier than before.
But this digitalisation will also lead to changes in the way societies
interact and therefore will translate into changes in how academia perceives itself.
We have also not sufficiently delved into the interaction with industry.
There are also some trends that we believe will stay with us over the coming years.
The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
Societies will be facing new questions, arising from technologies such as generally available AI systems.

We hope that the standard of research software will have risen across the board,
enabling easier reuse of prior software outputs in new research.
We also hope that a cultural change is taking place to recognize research software
as an integral part of science that it is widely recognised and rewarded across
academia and is supported by political actors.
\section{Conclusion}
We have tried to elaborate on the state on the future ten years from now.
It turned out to be a difficult endeavour, since even the full breadth of today's trends is not yet foreseeable.
We have tried to shed some light on the social, technical and organisational aspects.
In summary, RSEs will very likely be doing different work to what they are doing today,
but their core task of enhancing the scientific output of researchers -- other human
beings -- will stay the same.
They will still develop software artefacts together with scientists,
since better software leads to better research,
and better research leads to better software.

>>>>

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\author{
Florian Goth\authorOrcid{0000-0003-2707-4790}\authorRef{1},
Leyla Jael Castro\authorOrcid{0000-0003-3986-0510}\authorRef{2},
Gerasimos Chourdakis\authorOrcid{0000-0002-3977-1385}\authorRef{3},
\texorpdfstring{\\}{} % line break
Simon Christ\authorOrcid{0000-0002-5866-1472}\authorRef{4},
Jeremy Cohen\authorOrcid{0000-0003-4312-2537}\authorRef{5},
Nicholas A. Del Grosso\thinspace\authorRef{6},
\texorpdfstring{\\}{} % line break
Jean-Noël Grad\authorOrcid{0000-0002-5821-4912}\authorRef{7},
Magnus Hagdorn\authorOrcid{0000-0002-5076-4864}\authorRef{8},
Toby Hodges\authorOrcid{0000-0003-1766-456X}\authorRef{9},
Jan Linxweiler\authorOrcid{0000-0002-2755-5087}\authorRef{10},
\texorpdfstring{\\}{} % line break
Frank Löffler\authorOrcid{0000-0001-6643-6323}\authorRef{11},
Michele Martone\authorOrcid{0000-0003-3239-8554}\authorRef{12},
Jan Philipp Thiele\authorOrcid{0000-0002-8901-6660}\authorRef{10}\textsuperscript{,}\authorRef{13}
} % Authors and references to addresses
\institute{% Institutes with labels
\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\\
\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\\
\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\\
\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\\
\autlabel{5} Imperial College London, London, UK\\
\autlabel{6} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany\\
\autlabel{7} Institute for Computational Physics, University of Stuttgart, Germany\\
\autlabel{8} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\\
\autlabel{9} The Carpentries, USA\\
\autlabel{10} Technische Universität Braunschweig, Germany\\
\autlabel{11} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\\
\autlabel{12} Independent researcher, Germany\\
\autlabel{13} Weierstrass Institute, Berlin, Germany;
              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\\
}
\abstract{
It is 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSEng)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
As the field develops, the role and value of Research Software Engineers (RSEs) becomes clearer.
With the growing importance and scope of software engineering in research,
researchers who code now begin to fill dedicated, specialised RSE roles crucial for the sustainable development
of long-term research software projects that can rapidly innovate.
Looking into the future, participants of the deRSE25 conference discussed how RSEng could evolve in response to the
rapid changes in the broader social, technological, and organisational landscape.
This paper presents the main discussion points of the conference participants,
imagining a future in which RSEs have a well-defined and well-integrated role in research institutions,
working as engineers of ever more complex research software projects,
taming a growing range of methods and tools,
and guiding researchers through technical and ethical considerations
related to research software.
} % Abstract of the article
\keywords{Future RSE, Evolving field, Vision} % Keywords for the article

@@

\begin{story}{The Definition of RSEs}{defrse}
Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who transformed the chaos back into meaningful order.
They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
Their job titles have changed, but their role as bridges between logic and life remains the same.
In a future of shifting code and constant reinvention, it is their understanding of people that never went out of date.

@@

Further risks that were later discussed include an intensified replication crisis,
a de-democratization of research, and loss of trust in science.
Due to the non-deterministic nature of generative AI, we expect AI-assisted simulations
and LLM-assisted data analysis to become less reproducible.
This is not an entirely new issue, and similar concerns exist in the HPC community,
where reproducing simulation data and data analysis workflows presents tough challenges~\cite{Antunes2024}.
Professional RSEs could safeguard reproducibility by designing, evaluating, and monitoring appropriate workflows.
When applicable, standardised and version-controlled models should be preferred
over self-developed models, and self-developed models should be treated like any
other research output (i.e. publish the trained model, training data and training workflow).
Future research will require significant computing and storage resources, as well as useful data and models,
therefore it will be more likely to be driven by institutions that are able to provide these.
Finally, an increasing risk is the loss of trust in science,
in particular due to the automation of the peer-reviewing process~\cite{Naddaf2025}
and the malicious use of hidden prompts~\cite{Gibney2025}.
As researchers increase their reliance on AI assistants,
RSEs will be freed from repeatedly crafting similar individual elements,
but will be spending more of their time designing and reviewing software written by other humans and AI tools.
To address these risks, the RSE of the future should not only steer every part
of the research software development cycle (ensuring dissemination, archiving, and reproducibility),
but also contribute to endangered community resources and infrastructure that
will keep research accessible to the wider scientific community.

@@

% society implications
The work being undertaken by RSEs may well have direct, or indirect,
ethical implications in areas such as social engineering~\cite{s2erc,Siadati2024},
climate~\cite{Lannelongue2023} and more.
Like all areas of research and engineering RSEs are faced with the dual use dilemma~\cite{Bobier2024} where artefacts they develop can be used for military purposes.
As engineers, RSEs take the responsibility for the software they develop.
This includes an awareness of ethical issues that may arise.
RSEs need to be aware of these issues and develop their own ``red lines''
which need to be supported by organisational and professional guidelines.
With the advantages of \autoref{sec:ai}, ever-more AI created artefacts will be included in research.
As a result of this, we expect a discussion within societies about their acceptable use.

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{}{academic}
We expect research software to be a central part of scientific practice, on the same level as data and publications.
This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
We also hope that specialised clusters form, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
At universities which get such RSE units, we want to see that researchers naturally accept them as part of their available services.
Additionally, researchers will seamlessly interact with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of generations.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

@@

Besides all the improvements, the workshop participants still expect some challenges to persist,
and new challenges to emerge.
Due to the rise of dedicated RSE units, networking people and teams,
federating infrastructures, as well as organising and sharing knowledge will become important.
Structuring these new units and joining their administrative support will also help the units to grow.
A particular example is hiring:
While a standardised RSE Master's degree will make it easier to develop the next generation of RSEs,
organisations could develop standardised hiring processes to acquire the skill set that they actually need.
Budgeting and administrative hurdles will also probably still exist and hinder the formation of permanent and well-funded RSE teams.
However, we expect the growing (perceived) importance of research software
to convince research councils and funders to provide reliable streams of funding.

@@

\begin{story}{New career paths}{New career paths}
Ken, though trained as a physicist, found his passion not in theory,
but in keeping the quantum computers humming. He became the go-to expert
for troubleshooting entangled qubits and stabilising delicate cryogenic systems.
His unique blend of physics knowledge and hands-on problem solving earned him
a new title among his peers: the first Quantum RTP.
In a world where theory meets reality, Ken quietly ensures that the future
of quantum computing stays on track.
\end{story}
\section{What is in this future for RSEs?}
%FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
%GC-NOTE: Shall we add another story box here? Ideas what we write here?
In the previous sections, we have tried to list some of the trends that we foresee.
The material presented here is based on the outputs of a workshop, and extended with the authors' ideas.
Nevertheless, there are some things that we declared out of scope here.
Among that is the ongoing digital transformation in wider society.
Digital competencies in the general population continue to increase,
making the jump to software development easier than before.
But this digitalisation will also lead to changes in the way societies
interact and therefore will translate into changes in how academia perceives itself.
We have also not sufficiently delved into the interaction with industry.
There are also some trends that we believe will stay with us over the coming years.
The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
Societies will be facing new questions, arising from technologies such as generally available AI systems.

We hope that the standard of research software will have risen across the board,
enabling easier reuse of prior software outputs in new research.
We also hope that a cultural change is taking place to recognise research software
as an integral part of science that it is widely recognised and rewarded across
academia and is supported by political actors.
\section{Conclusion}
We have tried to elaborate on the state on the future ten years from now.
It turned out to be a difficult endeavour, since even the full breadth of today's trends is not yet foreseeable.
We have tried to shed some light on the social, technical and organisational aspects.
In summary, RSEs will very likely be doing different work to what they are doing today,
but their core task of enhancing the scientific output of researchers -- other human
beings -- will stay the same.
They will still develop software artefacts together with scientists,
since better software leads to better research,
and better research leads to better software.
